State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources

### 2021-03-29 analysis/paper/paper.Rmd
rk revisions to .Rmd and reviewer comments

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon and steelhead from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

Populations of Chinook Salmon *Oncorhynchus tshawytscha* and steelhead trout *O. mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric projects on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015]. As a result, most Snake River spring/summer-run Chinook Salmon (hereafter sp/sum Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River sp/sum Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River sp/sum Chinook Salmon and steelhead populations, with the exception of Tucannon River (a tributary of the Snake River whose confluence is downstream of Lower Granite Dam [LGR]), is monitored at LGR located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at LGR [@NMFS2019; @NWFSC2015; @NPCC2014]. As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, can be informed by escapements at LGR.

@@

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal. Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detection of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observation error variance (e.g., observation error at the window, or sampling variance at the trap). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish. Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for download or installation from [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated "truth". We then applied this model to sp/sum Chinook Salmon and steelhead returns at LGR for spawn years 2010 - 2019.

@@

Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at LGR, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at LGR and detected at arrays in the adult fish ladder. Below, we describe each of the data sources in more detail as they pertain to LGR; similar data could likely be obtained from other fish passage facilities.

@@

$$
Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right) \\
$$

The set trapping rate (i.e., the pre-determined time that the trap is open each hour to trap adults) does not always reflect the true trapping rate or proportion of fish that are actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error (pulses or lack thereof of passing adults), among others. Therefore, we use a mark-recapture approach to estimate a "true" trapping rate, $\nu_t$. The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGR that week, and using a Lincoln-Peterson mark-recapture model [@Seber2002].

$$
m_t \sim \text{Bin} \left( \nu_t, M_t \right) \\
$$

@@

We looked for differences in nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

@@

$$
\begin{aligned}
  \begin{bmatrix} X_{w,t} \\ X_{hc,t} \\ X_{hnc,t} \end{bmatrix}
  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{hc,t} \\ \omega_{hnc,t} \end{bmatrix}
\end{aligned}
$$

@@

Our objective was to assess STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement for accuracy and precision, given a known simulated "truth", under different possible conditions. We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios overall (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at LGR between 2010-2015. We generated 99 simulations for each scenario, and ran STADEM on each one, resulting in a total of 1,188 model runs. Further details about the simulation procedures can be found in [Appendix A](#append1).

To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures. Accuracy was captured by the relative bias: the difference between the simulated "truth" and the STADEM estimate, divided by the simulated "truth". We estimated the precision by examining the average coefficient of variation (CV) of the estimates. We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias. Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value. To compare the STADEM results directly to window counts, we compared the model estimates of total unique fish and the simulated window counts to the simulated number of unique fish.

@@

We applied STADEM to empirical data from LGR for sp/sum Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019. Window counts for both species were accessed from DART via functions within STADEM. For sp/sum Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past LGR starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past LGR between July 1, 2016 and June 30, 2017). Data from the adult trap were made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at LGR were accessed from the PTAGIS regional database ([https://www.ptagis.org/](https://www.ptagis.org/)).

@@

STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios as seen in the consistent interquartile distance in the boxplots of Figure \@ref(fig:rel-bias-lgd). For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R). In those scenarios, they were biased by nearly 5% (Figure \@ref(fig:rel-bias-lgd)), the difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)). The variance of the window counts also grew as the simulated observer error grew (no error vs. Err L vs. Err H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.

@@

We applied STADEM to data from LGR for sp/sum Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery clipped, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Estimates of total unique fish escaping past LGR were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years. Window count bias ranged from -3.0% to 11.0% for sp/sum Chinook Salmon and -9.9% to 6.6% for steelhead. Coefficients of variation ranged from 2.5 to 7.1% for wild fish, 2.3 to 5.1% for clipped hatchery fish, 3.4 to 8.8% for hatchery no-clip fish and 2.2 to 4.7% for total unique fish past LGR.

Weekly estimates of escapement over LGR tracked the window counts and trap estimates, with smaller credible intervals than the trap estimates alone. (One example: Figure \@ref(fig:jags-fit-fig). Similar plots for all model runs [each species and year combination] are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).

Weekly nighttime passage and re-ascension proportions were significantly different (2-sample test of proportions, *p* < 0.05) in approximately 50% of the weeks with sufficient PIT tag data to test. The differences in the estimated weekly rates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)). In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%.

@@

We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., LGR) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder. Our model explicitly accounts for nighttime passage, re-ascension, and potential error in both window and trap estimates. In doing so, we demonstrated that at LGR, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years. Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty. Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses [@Paulsen2007].

@@

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of sp/sum Chinook Salmon and steelhead past LGR, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020]. Estimates of species and origin-specific escapement at LGR, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM estimates of unique wild fish at LGR are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@Orme2019]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations. With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River sp/sum Chinook Salmon and steelhead groups [@IPTDSW2020].

@@

Funding for this study and development of the STADEM model was partially provided by the Bonneville Power Administration under project 2003-017-00. Special thanks to Darren Ogden and staff at the LGR adult trap for their hard work and diligent data collection. Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and to Matthew Campbell and staff at the Eagle Fish Genetics Laboratory for analyzing genetic samples. Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular. And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, as well as Eric Buhle, Sarah Hoffmann, John Hargrove and all the other people who have contributed through productive critique and conversations.

@@

```{r night-reasc-diff-fig, fig.height=5, fig.cap = "Nighttime passage rate plotted against re-ascension rate on the logit scale, calculated from observed PIT tags for each week of spawn years 2010-2019. The size of each point is proportional to the window count that week. The dashed line is the 1-1 line."}

>>>>

---
title: "State-Space Model to Estimate Salmon Escapement Using Multiple Data Sources"
author:
  - Kevin E. See:
      email: Kevin.See@merck.com
      institute: [BM]
      correspondence: true
  - Ryan N. Kinzer:
      email: ryank@nezperce.org
      institute: [NezPerce]
      correspondence: false
  - Michael W. Ackerman:
      email: mike.ackerman@merck.com
      institute: [BM]
      correspondence: false
institute:
  - BM: Biomark, Inc., 705 South 8th Street, Boise, Idaho, 83702, USA
  - NezPerce: Nez Perce Tribe, Department of Fisheries Resource Management, 14054 Burr Road, PO Box 1942, McCall, Idaho, 83638, USA
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
    bookdown::pdf_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      geometry: margin=1in
      latex_engine: "xelatex"
      fig_width: 7
      # fig_height: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
      includes:
        in_header: ../templates/header_manuscript.tex
    bookdown::html_document2:
      fig_caption: yes
      number_sections: FALSE
      toc: no
      fig_width: 7
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks2.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      number_sections: FALSE
      toc: no
      always_allow_html: true
      fig_width: 6
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
fontsize: 12pt
mainfont: Times New Roman
bibliography:
- references.bib
- packages.bib
csl: "../templates/american-fisheries-society.csl" # Insert path for the bib-style
abstract: |
  Accurate estimates of salmonids passing Lower Granite Dam on the Snake River, by species and origin, are a critical input to assessing the status and trends of various populations, as well as successful management of fisheries in the Snake River basin. Here we describe a state-space model that estimates such escapement past a dam by using window counts, passive integrated transponder (PIT) tag observations and data from an adult fish trap, accounting for issues such as nighttime passage, fallback and re-ascension, potential observation error at the window and uncertainty in the adult trap rate. We tested the approach using a simulation framework that mimicked several levels of observation error, differences between nighttime passage and re-ascension rates and the possibility of the adult trap being closed for some period of time. Our results demonstrate that the model produced unbiased estimates across all tested scenarios. We also applied this model to empirical data from the Lower Granite Dam to produce estimates of wild, clipped hatchery and unclipped hatchery spring/summer-run Chinook Salmon *Oncorhynchus tshawytscha* and steelhead *O. mykiss* from spawn years 2010-2019. 
keywords: |
  escapement; Chinook; steelhead; salmon; Lower Granite Dam; state-space model; Bayesian model
highlights: |
  These are the highlights. 
---

@@

Populations of Chinook Salmon *Oncorhynchus tshawytscha* and anadromous steelhead trout *O. mykiss* in the Snake River basin of Pacific Northwest, USA, are depleted following decades of substantial harvest and anthropogenic changes to their migration corridor (e.g., construction of hydroelectric projects on the Snake and Columbia rivers) and tributary habitats [@Nehlsen1991; @McClure2003; @NWFSC2015]. As a result, most Snake River spring/summer-run Chinook Salmon (hereafter sp/sum Chinook Salmon) were classified as threatened in 1992 under the Endangered Species Act (ESA; Federal Registry Notice 57 FR 14653), and Snake River summer-run steelhead trout (hereafter steelhead) were listed as threatened five years later (Federal Registry Notice 62 FR 43937). Snake River sp/sum Chinook Salmon and steelhead have substantial cultural, recreational, commercial and subsistence value within the Snake River basin as well as in downstream corridors (i.e., Columbia River) and ocean fisheries. The aggregate escapement of Snake River sp/sum Chinook Salmon and steelhead populations, with the exception of Tucannon River (a tributary of the Snake River whose confluence is downstream of Lower Granite Dam [LGR]), is monitored at LGR located in southeast Washington; the final dam on the Snake River that returning adults must pass prior to heading to tributary spawning locations. Many fisheries management and conservation actions are made based on estimates of escapement by species and origin at LGR [@NMFS2019; @NWFSC2015; @NPCC2014]. As an example, tribal ceremonial and subsistence harvest rates in the region are informed by escapement at LGR; additionally, the openings, closures and bag limits of fisheries targeting hatchery fish, both upstream in Snake River fisheries and downstream in mainstem Snake and Columbia rivers, are informed by escapements at LGR[@UnitedStatesOregon].

@@

In this study, we describe a novel method to estimate aggregate and group escapements which incorporate all sources of known uncertainty and demonstrate that the estimation method is essentially unbiased, thus better informing conservation and management decision making. We also test whether observed nighttime passage and fallback/re-ascension rates are typically equal. Our method for estimating species-specific escapement past LGR incorporates window counts, data from the adult fish trap located on the ladder, and passive detection of adults in the ladder from fish previously tagged with passive integrated transponder (PIT) tags to explicitly model nighttime passage, re-ascension, and observation error using a state-space approach [@Royle2008] which separates process variance (e.g., week to week variance in true escapement) from observational variance (e.g., observation error at the window, or sampling variance at the trap). To meet desired management and conservation objectives, modeled escapement includes estimates of uncertainty and is parsed into weekly strata. Further, total and weekly estimates are parsed into three origin groups: wild fish, hatchery fish with a clipped adipose fin, and unclipped hatchery fish. Our model is implemented in the **ST**ate-space **A**dult **D**am **E**scapement **M**odel (STADEM) package for the statistical software R [@R-Core-Team2020], and is available for download or installation from [https://github.com/KevinSee/STADEM](https://github.com/KevinSee/STADEM). To validate the STADEM results, we simulated 12 scenarios with varying trapping rates, fallback and re-ascension rates, nighttime passage rates, and window count error rates and compared STADEM results to the simulated "truth". Then, as a case study, we applied this model to sp/sum Chinook Salmon and steelhead returns at LGR for spawn years 2010 - 2019.

@@

Data sources for STADEM included 1) counts of fish migrating past the observation window located on the adult fish ladder at LGR, 2) information from adults captured at a fish trap located in the fish ladder, and 3) observations of fish PIT tagged prior to arriving at LGR and detected at arrays in the adult fish ladder. Below, we describe each of the data sources in more detail as they pertain to LGR; similar data could likely be obtained from other species passing LGR (e.g., fall Chinook Salmon) or other fish passage facilities (e.g., Bonneville Dam).

@@

$$
Y^T_t \sim \text{Bin} \left( \nu_t, X_t \right)
$$

The set trapping rate (i.e., the pre-determined time that the trap is open each hour to trap adults) does not always reflect the true trapping rate or proportion of fish actually captured in the trap due to various issues including trap malfunctions, separation-by-code fish opening the trap more frequently than expected, and process error (pulses or lack thereof of passing adults), among others. Therefore, we use a mark-recapture approach to estimate a "true" trapping rate, $\nu_t$. The estimate of the true weekly trap rate is derived based on previously PIT-tagged sp/sum Chinook Salmon and steelhead that are crossing LGR that week, and using a Lincoln-Peterson mark-recapture model [@Seber2002].

$$
m_t \sim \text{Bin} \left( \nu_t, M_t \right)
$$

@@

We looked for differences in weekly nighttime passage and re-ascension rates estimated using wild fish only, versus combining hatchery and wild fish together, and found no difference. Additionally, sample sizes of observed tags during most weeks are small, and may prohibit reliable estimation of rates for individual wild or hatchery stocks. Therefore, we combine wild and hatchery PIT tagged fish observations to estimate common nighttime passage and re-ascension rates to increase sample sizes.

@@

$$
\begin{aligned}
  \begin{bmatrix} X_{w,t} \\ X_{hc,t} \\ X_{hnc,t} \end{bmatrix}
  &= X_t * (1 - \eta_t) * \begin{bmatrix} \omega_{w,t} \\ \omega_{hc,t} \\ \omega_{hnc,t} \end{bmatrix}
\end{aligned}
$$
In practice, the total number of fish crossing LGR each week, $X_t$, could be split into other discrete groups if desired; for example, individual wild stocks using genetic stock identification or hatchery stocks with parentage based tagging [@Steele2019].

@@

Our objective was to assess STADEM model estimates of origin-specific (wild, hatchery, hatchery no-clip) escapement for accuracy and precision, given a known simulated "truth", under different possible conditions. We developed those conditions from the combinations of two trap rate scenarios (constant and shut down for 3 weeks), two fallback/re-ascension and nighttime passage combinations (equal and unequal), and three window count error rates (none, low and high); resulting in twelve different scenarios overall (Table \@ref(tab:lgr-sim-tab)). The simulation parameters such as proportion of origin, run-timing, nighttime passage rates, fallback and re-ascension rates and trap rates were based on observed values at LGR between 2010-2015. We generated 500 simulations for each scenario, and ran STADEM on each one, resulting in a total of 6,000 model runs. Further details about the simulation procedures can be found in [Appendix A](#append1).

To assess the performance of the model, we evaluated the estimates of origin-specific escapement by several measures. Accuracy was captured by the relative bias: the difference between the simulated "truth" and the mean STADEM estimate for each scenario, divided by the simulated "truth". We estimated the precision by examining the average coefficient of variation (CV) of the estimates. We calculated the root mean squared error (RMSE) as the square root of the mean of the squared bias in the estimate, which is equivalent to the square root of the variance of the estimator plus the square of the expected bias. Finally, we evaluated whether estimates of uncertainty were appropriate by calculating coverage probabilities by determining what proportion of the model results generated an estimated 95% credible interval that contained the simulated true value. To compare the STADEM results directly to window counts, we compared the model estimates of total unique fish and the simulated window counts to the simulated number of unique fish.

@@

We applied STADEM to empirical data from LGR for sp/sum Chinook Salmon and steelhead returning to the Snake River during spawn years 2010 to 2019. Window counts for both species were accessed from DART via functions within STADEM. For sp/sum Chinook Salmon, a spawn year refers to adults that migrate past the dam prior to August 17 each year and spawn that late summer and fall. For steelhead, the spawn year is defined as steelhead that migrate past LGR starting July 1 the previous year and prior to June 30 of the given year (e.g., spawn year 2017 steelhead migrate past LGR between July 1, 2016 and June 30, 2017). Data from the adult trap were made available by Idaho Department of Fish and Game, and adult PIT tag detection data within the fish passage ladder at LGR were accessed through DART and the adult ladder PIT tag query ([http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail](http://www.cbr.washington.edu/dart/query/pitadult_obsyr_detail)).

@@

STADEM results showed similar patterns for estimates of unique total fish and unique wild fish, in that both were unbiased with similar variance in the estimator across all scenarios as seen in the consistent interquartile range in the boxplots of Figure \@ref(fig:rel-bias-lgd). For unique total fish, window counts were unbiased, except in scenarios when night passage and re-ascension rates were not equal (N-R). In those scenarios, they were biased by nearly 5% (Figure \@ref(fig:rel-bias-lgd)), the designed simulated difference between night passage and re-ascension rates (Table \@ref(tab:lgr-sim-tab)). The variance of the window counts also grew as the simulated observer error grew (no error vs. Err L vs. Err H), whereas the variance in the STADEM estimates remained fairly constant, albeit often larger than the window count variance.

@@

We applied STADEM to data from LGR for sp/sum Chinook Salmon and steelhead for spawn years 2010–2019. Estimates of total escapement, as well as estimates of wild, hatchery clipped, and hatchery no-clip estimates are presented in Table \@ref(tab:lgr-tab). Estimates of total unique fish escaping past LGR were sometimes higher and sometimes lower than the raw window counts, indicating that the relative strength of nighttime passage and re-ascension rates differed across years. Window count bias ranged from -3.0% to 11.0% for sp/sum Chinook Salmon and -9.9% to 6.6% for steelhead. STADEM coefficients of variation ranged from 2.5 to 7.1% for wild fish, 2.3 to 5.1% for clipped hatchery fish, 3.4 to 8.8% for hatchery no-clip fish and 2.2 to 4.7% for total unique fish past LGR.

Weekly STADEM estimates of escapement over LGR tracked the window counts and trap estimates, with smaller credible intervals than the trap estimates alone. (One example: Figure \@ref(fig:jags-fit-fig). Similar plots for all model runs [each species and year combination] are available on the manuscript GitHub page: [https://www.github.com/KevinSee/ManuscriptSTADEM]([https://www.github.com/KevinSee/ManuscriptSTADEM])).

Weekly nighttime passage and re-ascension proportions were significantly different (2-sample test of proportions, *p* < 0.05) in approximately 50% of the weeks with sufficient PIT tag data to test. The differences in the estimated weekly rates were also apparent visually (Figure \@ref(fig:night-reasc-diff-fig)). In particular, there were several weeks when the window counts were quite large, but the nighttime passage and re-ascension rates differed by as much as 10%; leading to larger window count bias in some years \@ref(tab:lgr-tab).

@@

We have presented a novel method for estimating adult salmonid escapement past a large hydroelectric facility (e.g., LGR) that incorporates data from window counts, a fish trap, and observations of PIT tagged fish in the adult passage ladder. Our model explicitly accounts for nighttime passage, re-ascension, and potential error in both window and trap estimates. In doing so, we demonstrated that at LGR, nighttime passage and re-ascension rates do not always offset each other, and assuming they do will lead to biased escapement estimates in some years. Our state-space model combined multiple imperfect sources of data to reduce bias in adult escapement estimates and provided quantitative estimates of uncertainty. Using both data streams proved beneficial, and protected estimates from unforeseen problems or periodic biases in observations (e.g., trap closures, unequal night time and re-ascension rates). Accurate population or group abundance estimates and uncertainty accounting for observation and process error can be particularly important when estimates are used for management and conservation decisions such as population viability analyses [@Paulsen2007].

@@

Recently, co-managers in the Snake River basin have adopted the STADEM framework to estimate escapement of sp/sum Chinook Salmon and steelhead past LGR, and returning to tributary or population specific spawning areas [@IPTDSW2020; @Kinzer2020]. Estimates of species and origin-specific escapement at LGR, including known uncertainty, are available to further parse into sex- or age-structured escapement estimates [e.g., @Camacho2018; @Schrader2013] that are important for fisheries management and productivity monitoring of wild populations. As an example, STADEM estimates of unique wild fish at LGR are combined with estimated movement or transition probabilities based on PIT tag observations at instream PIT tag detection systems throughout the Snake River basin, similar to @Waterhouse2020, to estimate escapement to Snake River populations and locations throughout the basin [@IPTDSW2020; @Kinzer2020; @Orme2019]. Combined, escapement estimates from STADEM and movement probability estimates provide abundance estimates to given tributaries or populations. With sex and age data collected at the adult fish trap [@Hargrove2019], this approach provides necessary information to evaluate productivity and population viability for select Snake River sp/sum Chinook Salmon and steelhead groups [@IPTDSW2020].

@@

Funding for this study and development of the STADEM model was partially provided by the Bonneville Power Administration under project 2003-017-00. Special thanks to Darren Ogden and staff at the LGR adult trap for their hard work and diligent data collection. Thank you to personnel at the Idaho Department of Fish and Game, particularly Paul Bunn, Tim Copeland, and Bill Schrader, for providing access to data from the adult fish trap and conceptualizing methods, and to Matthew Campbell and staff at the Eagle Fish Genetics Laboratory for analyzing genetic samples. Special thanks to Columbia Basin Research staff and the Columbia River Data Access in Real Time (DART) application, and Susannah Iltis in particular. Thank you to three anonymous reviewers that helped us improve the manuscript with their thoughtful comments and suggestions. And finally, thank you to Rick Orme and Brice Semmens for their contributions to the development of STADEM, as well as Eric Buhle, Sarah Hoffmann, John Hargrove and all the other people who have contributed through productive critique and conversations.

@@

```{r night-reasc-diff-fig, fig.height=5, fig.cap = "Nighttime passage rate plotted against re-ascension rate on the logit scale, calculated from observed PIT tags for each week of spawn years 2010-2019. The size of each point is proportional to the window count that week. Points falling below the 1-1 dashed line indicate weekly periods of higher nighttime passage as compared to re-ascension, conversly, points falling above indicate a lower nighttime passage than re-ascension rate."}
